---
title: 'R-code for "Dancing in the moonlight: evidence that Killdeer foraging behavior varies
  with lunar cycle"'
author: "Luke J. Eberhart-Phillips"
output:
  pdf_document: 
    fig_height: 5
    fig_width: 5
  html_document:
    highlight: pygments
---
```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", cache = TRUE) # warning = FALSE
```

In this document I provide all the necessary code for repeating the analyses presented in my paper.  To access the dataset and Rmarkdown file, please download this [GitHub](https://github.com/leberhartphillips/killdeer_foraging) repository.  Simply follow the link and click on *Download ZIP* on the right-hand side of the page.  An explanation of the files in the repository can be found in the Readme file.  Please don't hesitate to contact me at `luke.eberhart[at]gmail.com` if you have any questions.

The structure of the code I present here follows the analyses presented in the *Results* section of the paper.

**Prerequisites:**

* For running the complete code you need a `files` subfolder with the raw data file found in the [GitHub](https://github.com/leberhartphillips/killdeer_foraging) repository.
* The following packages are needed for analysis and can be easily installed from [CRAN](http://cran.r-project.org/) by uncommenting the `install.packages` functions:

```{r, results="hide", message=FALSE, cache = TRUE}
# install.packages("dplyr")
# install.packages("gridExtra")
# install.packages("ggplot2")
# install.packages("RColorBrewer")
# install.packages("grid")
# install.packages("extrafont")
# install.packages("lme4")
# install.packages("AICcmodavg")
# install.packages("VGAM")
# install.packages("tidyr")
# install.packages("stringr")
# install.packages("scales")
# install.packages("multcomp")
# install.packages("Rmisc")
# install.packages("arm")
library(dplyr)
library(gridExtra)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(extrafont)
library(lme4)
library(AICcmodavg)
library(VGAM)
library(tidyr)
library(stringr)
library(scales)
library(multcomp)
library(Rmisc)
library(arm)
library(reshape)
```

***
## Loading and wrangling data

Loading the 

- **behavior data** in which each row is a single observation of the behavior (i.e. foraging, roosting, alert, or preening) of an individual on a given date, time, and scan. Within each behavior, a "1" signifies the behavior the individual was engaged in, whereas a "0" signifies the behavior the individual was not engaged in. 
- **lunar data** of the proportion of the moon's disk that was illuminated the preceding night of the date specified in each row.
- **climate data** for the maximum precipitation (mm) and minimum temperature (°C) recorded during the preceding 12 hours of each survey session.

```{r}
Behavior <- read.csv(".//Data_files//behavior_data_raw.csv")
Lunar <- read.csv(".//Data_files//lunar_data_raw.csv")
Climate <- read.csv(".//Data_files//climate_data_raw.csv")
```

Summarize the `Behavior` data such that each row reveals the total number of individuals engaged in a given behavior on each scan.

```{r}
Behavior_totals <- 
  Behavior %>%
  dplyr::group_by(Date, Time, Scan) %>%
  dplyr::summarise(Foraging = sum(Foraging),
                   Roosting = sum(Roosting),
                   Alert = sum(Alert),
                   Preening = sum(Preening))
```

Define the variables of the `Behavior_totals` correctly.  `Date` needs to be recognized as a date, `Time` needs to be a factor with the order: `Sunrise`, `Midday`, and `Sunset`, and the behaviors need to be numeric.

```{r}
Behavior_totals$Date <- as.Date(Behavior_totals$Date, "%Y-%m-%d")
Behavior_totals$Time <- factor(Behavior_totals$Time, 
                               levels = c("Sunrise", "Midday", "Sunset"))
Behavior_totals[,c("Foraging", "Roosting", "Alert", "Preening")] <- 
  lapply(Behavior_totals[,c("Foraging", "Roosting", "Alert", "Preening")], as.numeric)
```

Create a `Total` variable in the dataframe which sums the total number of birds observed on each scan

```{r}
Behavior_totals$Total <- rowSums(Behavior_totals[,c(4:7)])
```

Define the variables of the `Climate` dataframe correctly.  `Date` needs to be recognized as a date, `Time` needs to be a factor with the order: `Sunrise`, `Midday`, and `Sunset`, and `Min12Temp` and `Max12Precip` need to be numeric.

```{r}
Climate$Date <- as.Date(Climate$Date, "%Y-%m-%d")
Climate[,c("Min12Temp", "Max12Precip")] <- 
  lapply(Climate[,c("Min12Temp", "Max12Precip")], as.numeric)
Climate$Time <- factor(Climate$Time,
                       levels = c("Sunrise", "Midday", "Sunset"))
```

Define the variables of the `Lunar` dataframe correctly.  `Date` needs to be recognized as a date and `PropMoon` need to be numeric.

```{r}
Lunar$Date <- as.Date(Lunar$Date, "%Y-%m-%d")
Lunar[,c("PropMoon")] <- as.numeric(Lunar[,c("PropMoon")])
```

Join the three dataframes by `Date` and `Time` to create a collated dataframe henceforth referred to as `Killdeer`.  Order the data according to `Date`, `Time`, and `Scan`.

```{r}
Killdeer <- dplyr::left_join(Behavior_totals, Climate, by = c("Date", "Time"))
Killdeer <- dplyr::left_join(Killdeer, Lunar, by = c("Date"))
Killdeer <- Killdeer[with(Killdeer, order(Date, Time, Scan)), ]
head(Killdeer)
```

Subset dataset to include only scans that had at least one bird observed.

```{r}
Killdeer <- dplyr::filter(Killdeer, Total != 0)
```

Create a `Session` variable that defines each unique visit to the study site

```{r}
Killdeer$Session <- as.factor(paste(Killdeer$Date, Killdeer$Time, sep = "_"))
```

Create a `Unique_scan` variable that defines each unique scan made

```{r}
Killdeer$Unique_scan <- as.factor(paste(Killdeer$Session, Killdeer$Scan, sep = "_"))
```

Add variables describing the number of individuals not engaged in the focal behavior

```{r}
Killdeer$NonForaging <- Killdeer$Total - Killdeer$Foraging
Killdeer$NonRoosting <- Killdeer$Total - Killdeer$Roosting 
Killdeer$NonPreening <- Killdeer$Total - Killdeer$Preening 
Killdeer$NonAlert <- Killdeer$Total - Killdeer$Alert
```

Check sample sizes for `Date`, `Session`, and `Unique_scan`

```{r}
length(levels(as.factor(Killdeer$Date)))
length(levels(as.factor(Killdeer$Session)))
length(levels(as.factor(Killdeer$Unique_scan)))
```

***
## Temporal variation in killdeer abundance

Linear regression mixed model predicting variation in abundance across time of day (i.e. fixed effect: `Time`, random effects: `Unique_scan`, `Session`, `Date`).

```{r}
Abundance_Time_model <- 
  lme4::glmer(Total ~ 
              Time + (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = poisson)
```

Calculate the post hoc Tukey contrast statistics

```{r}
summary(multcomp::glht(Abundance_Time_model, mcp(Time = "Tukey")))
```

Calcluate summary statistics of abundance across the three times of day, then plot (Figure 2 in the paper).

```{r, fig.align='center'}
Abundance <- Rmisc::summarySE(Killdeer, measurevar = "Total", groupvars = c("Time"))
ggplot2::ggplot(data = Abundance, aes(x = Time, y = Total)) +
          geom_bar(position = position_dodge(), stat = "identity", alpha = 0.5) +
          geom_errorbar(aes(ymin = Total - ci, ymax = Total + ci), 
                        width = .2, position = position_dodge(.9)) +
          theme_bw() +
          theme(axis.title.x = element_text(size = 14, vjust = -0.5),
                axis.text.x  = element_text(size = 13), 
                axis.title.y = element_text(size = 14, vjust = 1.5),
                axis.text.y  = element_text(size = 13), 
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank()) +
          ylab("Abundance (± 95% CI)") +
          xlab("Time of survey")
```

***
## Foraging model selection, analysis, and plotting

Assign the *a priori* models to a list

```{r}
Forage_models <- list()
```

- **Model 1: Null model**

```{r, warning=FALSE}
Forage_models[[1]] <- 
  lme4::glmer(cbind(Foraging, NonForaging) ~ 
                1 + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 2: Univariate model with proportion of moon illuminated**

```{r, warning=FALSE}
Forage_models[[2]] <- 
  lme4::glmer(cbind(Foraging, NonForaging) ~ 
                PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 3: Univariate model with minimum 12 hour temperature**

```{r, warning=FALSE}
Forage_models[[3]] <- 
  lme4::glmer(cbind(Foraging, NonForaging) ~ 
                Min12Temp + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 4: Univariate model with max 12 hour precipitation**

```{r, warning=FALSE}
Forage_models[[4]] <- 
  lme4::glmer(cbind(Foraging, NonForaging) ~ 
                Max12Precip + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 5: Additive model of both lunar and temperature covariates**

```{r, warning=FALSE}
Forage_models[[5]] <- 
  lme4::glmer(cbind(Foraging, NonForaging) ~ 
                Min12Temp + PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 6: Additive model of both lunar and precipitation covariates**

```{r, warning=FALSE}
Forage_models[[6]] <- 
  lme4::glmer(cbind(Foraging, NonForaging) ~ 
                Max12Precip + PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 7: Additive model of both lunar, temperature, and precipitation covariates**

```{r, warning=FALSE}
Forage_models[[7]] <- 
  lme4::glmer(cbind(Foraging, NonForaging) ~ 
                Min12Temp + Max12Precip + PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

Name models and create a ranked AIC table

```{r}
Modnames <- paste("Model", 1:length(Forage_models), sep = " ")
Forage_models_table <- aictab(cand.set = Forage_models, modnames = Modnames, sort = TRUE)
```

Calculate evidence ratios and add them to the AIC table

```{r}
Evidence_Ratios <- function(AIC_table){
  ER <- vector( , nrow(AIC_table))
  for(i in 1:nrow(AIC_table)) {
    ER[i] <- AIC_table$AICcWt[1]/AIC_table$AICcWt[i]
  }
  ER[1] <- NA
  table <- cbind(AIC_table, ER)
  out <- cbind(table[, c(1:2)], round(table[, c(3:9)], 2))
}

Forage_models_table <- Evidence_Ratios(Forage_models_table)
```

Estimate the marginal R-squared of each model following Nakagawa and Schielzeth (Equ. 29 and 30 and Table 2; 2013) and add them to the AIC table to produce Table 1 presented in the paper.

```{r, warning=FALSE, message=FALSE}
r2_m_glmm <- function(mod){
  v <- var(as.vector(fixef(mod) %*% t(model.matrix(mod))))
  r2_m <- v/(v + VarCorr(mod)$Unique_scan[1] + 
               VarCorr(mod)$Session[1] + VarCorr(mod)$Date[1] + pi^2/3)
  out <- unlist(list(r2_m = r2_m))
}
r2_m <- data.frame(round(t(as.data.frame(t(sapply(Forage_models, r2_m_glmm)))), 3))
r2_m$Modnames <- 
  c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", "Model 6", "Model 7")
colnames(r2_m) <- c("r2_m", "Modnames")
Forage_models_table <- left_join(Forage_models_table, r2_m)[,c(1,2,4,6,9,10)]
Forage_models_table
```

Calculate the 95% confidence interval of the top model by creating a function which applies varying steps of the predictor (i.e., PropMoon) to the model and returns the coefficents of the intercept and beta.

```{r}
CI_foraging <- function(offs) {
  model <- glmer(cbind(Foraging, NonForaging) ~ 
              I(PropMoon-offs) + (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
  ests <- summary(model)$coefficients[1,1:2]
  # backlink the coefficients to the probability scale
  return(c(offs,ests,invlogit(ests[1]+c(-1,0,1)*1.96*ests[2])))
}
# specify the offs (i.e., vector of numbers from 0 to 1 stepped by 0.05)
offs_foraging <- seq(0,1,0.05)
# apply the offs vector to the function (retuning a matrix)
result_foraging <- sapply(offs_foraging,CI_foraging)
# transpose the matrix
result_foraging <- t(result_foraging)
# convert the matrix to a data.frame
result_foraging <- data.frame(result_foraging)
# define the column names
colnames(result_foraging) <- 
  c("PropMoon", "Coefficient", "Std. Error", "Upper", "Mean", "Lower")
# Create proportional variable for plotting
Killdeer$prop_Forage <- Killdeer$Foraging/Killdeer$Total
```

Plot the top model with confidence intervals

```{r, fig.align='center', fig.width=6}
ggplot2::ggplot() + 
          geom_point(data = Killdeer, 
                     aes(x = PropMoon, y = prop_Forage), size = 3, alpha = 0.5) + 
          geom_ribbon(data = result_foraging, 
                      aes(x = PropMoon, y = Mean, ymin = Lower, ymax = Upper), 
                      fill = "#1B9E77", alpha = 0.25) +
          geom_line(data = result_foraging, 
                    aes(x = PropMoon, y = Mean), size = 1.5, colour = "#1B9E77") +
          theme_bw() +
          theme(legend.position = c(1, 1), 
                legend.justification = c(1, 1),
                legend.text=element_text(size=11),
                legend.title=element_blank(),
                legend.key.height=unit(0.8,"line"),
                legend.key.width=unit(0.8,"line"),
                legend.background = element_rect(fill=NA),
                axis.title.x = element_text(size = 14, vjust = -0.5),
                axis.text.x  = element_text(size = 13), 
                axis.title.y = element_text(size = 14, vjust = 1.5),
                axis.text.y  = element_text(size = 13), 
                panel.grid.major = element_blank()) +
          scale_y_continuous(limits=c(0,1)) +
          ylab("Proportion of flock foraging in daylight") +
          xlab("Proportion of moon illuminated the previous night")
```
***
## Roosting model selection, analysis, and plotting

Assign the *a priori* models to a list

```{r}
Roosting_models <- list()
```

- **Model 1: Null model**

```{r, warning=FALSE}
Roosting_models[[1]] <- 
  lme4::glmer(cbind(Roosting, NonRoosting) ~ 
                1 + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 2: Univariate model with proportion of moon illuminated**

```{r, warning=FALSE}
Roosting_models[[2]] <- 
        lme4::glmer(cbind(Roosting, NonRoosting) ~ 
                PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 3: Univariate model with minimum 12 hour temperature**

```{r, warning=FALSE}
Roosting_models[[3]] <- 
  lme4::glmer(cbind(Roosting, NonRoosting) ~ 
                Min12Temp + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 4: Univariate model with max 12 hour precipitation**

```{r, warning=FALSE}
Roosting_models[[4]] <- 
  lme4::glmer(cbind(Roosting, NonRoosting) ~ 
          Max12Precip + 
          (1| Unique_scan) + (1 | Session) + (1| Date), 
        data = Killdeer, family = binomial)
```

- **Model 5: Additive model of both lunar and temperature covariates**

```{r, warning=FALSE}
Roosting_models[[5]] <- 
  lme4::glmer(cbind(Roosting, NonRoosting) ~ 
                Min12Temp + PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 6: Additive model of both lunar and precipitation covariates**

```{r, warning=FALSE}
Roosting_models[[6]] <- 
  lme4::glmer(cbind(Roosting, NonRoosting) ~ 
                Max12Precip + PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 7: Additive model of both lunar, temperature, and precipitation covariates**

```{r, warning=FALSE}
Roosting_models[[7]] <- 
  lme4::glmer(cbind(Roosting, NonRoosting) ~ 
          Min12Temp + Max12Precip + PropMoon + 
          (1| Unique_scan) + (1 | Session) + (1| Date), 
        data = Killdeer, family = binomial)
```

Name models, create a ranked AIC table, calculate evidence ratios and marginal r-squared, and produce Table 2 presented in the paper.

```{r, warning=FALSE, message=FALSE}
Modnames <- paste("Model", 1:length(Roosting_models), sep = " ")
Roost_models_table <- aictab(cand.set = Roosting_models, modnames = Modnames, sort = TRUE)
Roost_models_table <- Evidence_Ratios(Roost_models_table)
r2_m <- data.frame(round(t(as.data.frame(t(sapply(Roosting_models, r2_m_glmm)))), 3))
r2_m$Modnames <- 
  c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", "Model 6", "Model 7")
colnames(r2_m) <- c("r2_m", "Modnames")
Roost_models_table <- left_join(Roost_models_table, r2_m)[,c(1,2,4,6,9,10)]
Roost_models_table
```

Plot the top model with confidence intervals

```{r, fig.align='center', fig.width=6}
CI_roosting <- function(offs) {
  model <- glmer(cbind(Roosting, NonRoosting) ~ 
              I(PropMoon-offs) + (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
  ests <- summary(model)$coefficients[1,1:2]
  return(c(offs,ests,invlogit(ests[1]+c(-1,0,1)*1.96*ests[2])))
}
offs_roosting <- seq(0,1,0.05)
result_roosting <- sapply(offs_roosting,CI_roosting)
result_roosting <- t(result_roosting)
result_roosting <- data.frame(result_roosting)
colnames(result_roosting) <- 
  c("PropMoon", "Coefficient", "Std. Error", "Upper", "Mean", "Lower")
Killdeer$prop_Roost <- Killdeer$Roosting/Killdeer$Total
ggplot2::ggplot() + 
          geom_point(data = Killdeer, 
                     aes(x = PropMoon, y = prop_Roost), size = 3, alpha = 0.5) + 
          geom_ribbon(data = result_roosting, 
                      aes(x = PropMoon, y = Mean, ymin = Lower, ymax = Upper), 
                      fill = "#D95F02", alpha = 0.25) +
          geom_line(data = result_roosting, 
                    aes(x = PropMoon, y = Mean), size = 1.5, colour = "#D95F02") +
          theme_bw() +
          theme(legend.position = c(1, 1), 
                legend.justification = c(1, 1),
                legend.text=element_text(size=11),
                legend.title=element_blank(),
                legend.key.height=unit(0.8,"line"),
                legend.key.width=unit(0.8,"line"),
                legend.background = element_rect(fill=NA),
                axis.title.x = element_text(size = 14, vjust = -0.5),
                axis.text.x  = element_text(size = 13), 
                axis.title.y = element_text(size = 14, vjust = 1.5),
                axis.text.y  = element_text(size = 13), 
                panel.grid.major = element_blank()) +
          scale_y_continuous(limits=c(0,1)) +
          ylab("Proportion of flock roosting in daylight") +
          xlab("Proportion of moon illuminated the previous night")
```

***
## Preening model selection, analysis, and plotting

Assign the *a priori* models to a list

```{r}
Preening_models <- list()
```

- **Model 1: Null model**

```{r, warning=FALSE}
Preening_models[[1]] <- 
  lme4::glmer(cbind(Preening, NonPreening) ~ 
                1 + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 2: Univariate model with proportion of moon illuminated**

```{r, warning=FALSE}
Preening_models[[2]] <- 
  lme4::glmer(cbind(Preening, NonPreening) ~ 
                PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)

```

- **Model 3: Univariate model with minimum 12 hour temperature**

```{r, warning=FALSE}
Preening_models[[3]] <- 
  lme4::glmer(cbind(Preening, NonPreening) ~ 
          Min12Temp + 
          (1| Unique_scan) + (1 | Session) + (1| Date), 
        data = Killdeer, family = binomial)
```

- **Model 4: Univariate model with max 12 hour precipitation**

```{r, warning=FALSE}
Preening_models[[4]] <- 
  lme4::glmer(cbind(Preening, NonPreening) ~ 
                Max12Precip + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 5: Additive model of both lunar and temperature covariates**

```{r, warning=FALSE}
Preening_models[[5]] <- 
  lme4::glmer(cbind(Preening, NonPreening) ~ 
                Min12Temp + PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 6: Additive model of both lunar and precipitation covariates**

```{r, warning=FALSE}
Preening_models[[6]] <- 
  lme4::glmer(cbind(Preening, NonPreening) ~ 
                Max12Precip + PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

- **Model 7: Additive model of both lunar, temperature, and precipitation covariates**

```{r, warning=FALSE}
Preening_models[[7]] <- 
  lme4::glmer(cbind(Preening, NonPreening) ~ 
                Min12Temp + Max12Precip + PropMoon + 
                (1| Unique_scan) + (1 | Session) + (1| Date), 
              data = Killdeer, family = binomial)
```

Name models, create a ranked AIC table, calculate evidence ratios and marginal r-squared, and produce Table 3 presented in the paper.

```{r, warning=FALSE, message=FALSE}
Modnames <- paste("Model", 1:length(Preening_models), sep = " ")
Preen_models_table <- aictab(cand.set = Preening_models, modnames = Modnames, sort = TRUE)
Preen_models_table <- Evidence_Ratios(Preen_models_table)
r2_m <- data.frame(round(t(as.data.frame(t(sapply(Preening_models, r2_m_glmm)))), 3))
r2_m$Modnames <- 
  c("Model 1", "Model 2", "Model 3", "Model 4", "Model 5", "Model 6", "Model 7")
colnames(r2_m) <- c("r2_m", "Modnames")
Preen_models_table <- left_join(Preen_models_table, r2_m)[,c(1,2,4,6,9,10)]
Preen_models_table
```



###### This is a test ########